Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations: the Oziris algorithm

### 2016-11-01 ParisEM.tex
Minor modifications

\title{Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations}

@@

\section{Introduction}
It is assumed that the dynamics of the states $(X_t)_{t\ge 0}$ depends on a parameter $\theta\in \Theta$ where $\Theta$ is a compact subspace of $\mathbb{R}^{\ell}$ and that $(X_t)_{t\ge 0}$ is a weak solution to the following Stochastic Differential Equation (SDE) in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha_{\theta}(X_t)\rmd t + \rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion. In the following, it is assumed that $\alpha_{\theta}$ is of the form $\alpha_{\theta}(x) = \nabla_x A_{\theta}(x)$ where $A_{\theta}: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function. A popular choice of $A_{\theta}$  is $A_{\theta} = \log \pi_{\theta}/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi_{\theta}$. See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi_{\theta}$. The states $(X_t)_{t\ge 0}$ are only partially observed at discrete times $(t_k)_{0\le k \le n}$, with $t_0=0$, through the sequence $(Y_k)_{0\le k\le n}$ which is such that $(Y_k)_{0\le k\le n}$ are independent conditional on $(X_t)_{t\ge 0}$ and for all $0\le k\le n$, the conditional distribution of $Y_k$ given $(X_t)_{t\ge 0}$ depends only on $X_k = X_{t_k}$. This distribution has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g_{\theta}(X_k,\cdot)$. For all $1\le k \le n$, the distribution of $X_{k} $ conditional on $X_{k-1}$ has a density $q^{\Delta t_k}_{\theta}(X_k,\cdot)$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$.

Statistical inference for partially observed stochastic differential equations usually involves the computation of conditional distributions of some unobserved states given the observations $(Y_0,\ldots,Y_n)$. These posterior distributions are crucial to compute maximum likelihood estimators defined as maximizers of the incomplete data log-likelihood $\theta\mapsto \ell_{\theta}^{n}$ given by
\begin{equation*}
\ell_{\theta}^{n}(Y_{0:n}) = \log\left(\int p_{\theta}(x_{0:n},Y_{0:n})\,\rmd x_{0:n}\right)\eqsp,
\end{equation*}  
where the complete data likelihood $p_{\theta}$ is given by
\begin{equation*}
p_{\theta}(x_{0:n},Y_{0:n}) = \chi_{\theta}(x_0)g_{\theta}(x_0;Y_0)\prod^{n}_{k=1}q^{\Delta t_k}_{\theta}(x_{k-1},x_k)g_{\theta}(x_k,Y_k)\eqsp.
\end{equation*}
For instance, the EM algorithm introduced in \cite{dempster:laird:rubin:1977} iteratively builds a sequence $\{\theta_{p}\}_{p\ge 0}$ of parameter estimates following the two steps:
\begin{enumerate}
	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$, where $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$ is the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$.
	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
\end{enumerate}
The E-step boils down to the computation of the smoothed additive function: 
\begin{multline*}
Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi_{\theta}(X_0)g_{\theta}(X_0;Y_0)\right)\middle|Y_{0:n}\right] \\
+ \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left(q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)\right)\middle|Y_{0:n}\right] \eqsp.
\end{multline*}
Similarly, by Fisher's identity, recursive maximum likelihood estimates may be computed using the gradient of the loglikelihood which can be written as the conditional expectation of an additive functional of the hidden states given all the observations. See \cite[Chapter $10$ and $11$]{cappe:moulines:ryden:2005}, \cite{kantas:doucet:signh:2015,doucet:poyiadjis:singh:2011,lecorff:fort:2013a,lecorff:fort:2013b}
for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.

@@

\section{Random weight PaRIS algorithm}
%For a given value $\theta_p$ of the parameter, dropping the initial state, the E-step may be written:
%\[
%Q(\theta,\theta_{p})= \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left\{q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)\right\}\middle|Y_{0:n}\right]\eqsp.
%\]
%We first describe the Monte Carlo method used to approximate $Q(\theta,\theta_{p})$ when $q^{\Delta t_k}_{\theta}$ may be computed. Dropping $\theta$ from the notations, the E-step computes an expectation of the form
The aim of this Section is to detail the extension of PaRIS algorithm to approximate expectations of the form
\begin{equation}
\label{def:addfunc}
H_{n} =  \sum_{k=1}^n\mathbb{E}\left[h_k(X_{k-1},X_k)\middle|Y_{0:n}\right]\eqsp,
\end{equation}
when the transition density of the hidden states is not available explicitly where $\{h_k\}_{k=1}^n$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$. Thus, for $0 \leq k \leq k' \leq n$ and a given measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, we need to compute an approximation of the smoothed expectation of the states $(X_k,\ldots,X_{k'})$ defined by:
\[
\phi_{k:k'|n}[h] := \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
\]
The filtering distributions are defined, for $0\le k\le n$ by $\phi_{k} = \phi_{k:k|k}$. %In this paper, we are interested in a particular class of Sequential Monte Carlo methods to approximate $\bar{H}_{n}$ known as the PaRIS algorithm introduced in \cite{olsson:westerborn:2016}.

>>>>

\title{Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations: the Oziris algorithm}

@@

\section{Introduction}
The dynamics of the states $(X_t)_{t\ge 0}$ depends on a parameter $\theta\in \Theta$ where $\Theta$ is a compact subspace of $\mathbb{R}^{\ell}$ and that $(X_t)_{t\ge 0}$ is a weak solution to the following Stochastic Differential Equation (SDE) in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha_{\theta}(X_t)\rmd t + \rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion. In the following, it is assumed that $\alpha_{\theta}$ is of the form $\alpha_{\theta}(x) = \nabla_x A_{\theta}(x)$ where $A_{\theta}: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function. A popular choice of $A_{\theta}$  is $A_{\theta} = \log \pi_{\theta}/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi_{\theta}$. See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi_{\theta}$. The states $(X_t)_{t\ge 0}$ are only partially observed at discrete times $(t_k)_{0\le k \le n}$, with $t_0=0$, through the sequence $(Y_k)_{0\le k\le n}$ which is such that $(Y_k)_{0\le k\le n}$ are independent conditional on $(X_t)_{t\ge 0}$ and for all $0\le k\le n$, the conditional distribution of $Y_k$ given $(X_t)_{t\ge 0}$ depends only on $X_k = X_{t_k}$. This distribution has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g_{\theta}(X_k,\cdot)$. For all $1\le k \le n$, the distribution of $X_{k} $ conditional on $X_{k-1}$ has a density $q^{\Delta t_k}_{\theta}(X_k,\cdot)$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$.

Statistical inference for partially observed stochastic differential equations usually involves the computation of conditional distributions of some unobserved states given the observations $(Y_0,\ldots,Y_n)$. These posterior distributions are crucial to compute maximum likelihood estimators defined as maximizers of the incomplete data log-likelihood $\theta\mapsto \ell_{\theta}^{n}$ given by
\begin{equation*}
\ell_{\theta}^{n}(Y_{0:n}) = \log\left(\int p_{\theta}(x_{0:n},Y_{0:n})\,\rmd x_{0:n}\right)\eqsp,
\end{equation*}  
where $p_{\theta}$ is the complete data likelihood:
\begin{equation*}
p_{\theta}(x_{0:n},Y_{0:n}) = \chi_{\theta}(x_0)g_{\theta}(x_0;Y_0)\prod^{n}_{k=1}q^{\Delta t_k}_{\theta}(x_{k-1},x_k)g_{\theta}(x_k,Y_k)\eqsp.
\end{equation*}
For instance, the EM algorithm introduced in \cite{dempster:laird:rubin:1977} iteratively builds a sequence $\{\theta_{p}\}_{p\ge 0}$ of parameter estimates following the two steps:
\begin{enumerate}
	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$, where $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$ is the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$.
	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
\end{enumerate}
The E-step boils down to the computation of the smoothed additive function: 
\begin{multline*}
Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi_{\theta}(X_0)g_{\theta}(X_0;Y_0)\right)\middle|Y_{0:n}\right] \\
+ \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left(q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)\right)\middle|Y_{0:n}\right] \eqsp.
\end{multline*}
Similarly, by Fisher's identity, recursive maximum likelihood estimates may be computed using the gradient of the loglikelihood which can be written as the conditional expectation of an additive functional of the hidden states given all the observations. See \cite[Chapter $10$ and $11$]{cappe:moulines:ryden:2005}, \cite{kantas:doucet:signh:2015,doucet:poyiadjis:singh:2011,lecorff:fort:2013a,lecorff:fort:2013b}
for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.

@@

\section{Random weight PaRIS algorithm}
%For a given value $\theta_p$ of the parameter, dropping the initial state, the E-step may be written:
%\[
%Q(\theta,\theta_{p})= \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left\{q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)\right\}\middle|Y_{0:n}\right]\eqsp.
%\]
%We first describe the Monte Carlo method used to approximate $Q(\theta,\theta_{p})$ when $q^{\Delta t_k}_{\theta}$ may be computed. Dropping $\theta$ from the notations, the E-step computes an expectation of the form
Let $0 \leq k \leq k' \leq n$. The joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
\[
\phi_{k:k'|n}[h] := \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
\]
For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denotes the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
\begin{equation}
\label{def:addfunc}
H_{n} =  \sum_{k=1}^n\mathbb{E}\left[h_k(X_{k-1},X_k)\middle|Y_{0:n}\right]\eqsp,
\end{equation}
when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=1}^n$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$.  %In this paper, we are interested in a particular class of Sequential Monte Carlo methods to approximate $\bar{H}_{n}$ known as the PaRIS algorithm introduced in \cite{olsson:westerborn:2016}.
